Explore the SingleR results

### 2025-07-10 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/explore-singler-results.Rmd
WIP: dotplot customization 2

```{r}
# get top 7 upregulated marker genes per cell type
nbatlas_markers_df <- readr::read_tsv("../references/nbatlas-marker-genes.tsv") |>
  dplyr::filter(direction == "up") |>
  dplyr::group_by(NBAtlas_label) |>
  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
  dplyr::ungroup() |>
  dplyr::filter(rank_lfc <= 5) |>
  harmonize_celltypes(NBAtlas_label, label_recoded) |>
  dplyr::select(label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
nbatlas_markers_df

all_markers
```

@@

# get markers that are expressed
expressed_markers <- intersect(all_markers, expressed_genes)
expressed_markers
```

@@

# Join with cell type results and marker gene info
all_info_df <- celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler") |>
  dplyr::select(cell_id, label_recoded) |>
  dplyr::left_join(gene_exp_df, by = "cell_id") |>
  # account for the same gene being present in multiple cell types
  dplyr::left_join(nbatlas_markers_df, by = c("label_recoded", "ensembl_gene_id"), relationship = "many-to-many")
```

```{r}
marker_gene_dotplot(
  all_info_df,
  markers_df = nbatlas_markers_df,
  celltype_colors = celltype_pal
)
```

CODE WORKS, NEEDS HELP THOUGH!
what I need to do when I come back to this update my names to match allys code
- broad_celltype_group is her singler
- validation_group_annotation is her nbatlas marker gene df label

Then i can use the same code to make a consensus dotplot.

```{r}
min_cells <- 0
# prep for plots
# get total number of cells per final annotation group
total_cells_df <- all_info_df |>
  dplyr::select(cell_id, label_recoded) |>
  dplyr::distinct() |>
  dplyr::count(label_recoded, name = "total_cells")

# table with one row per unique broad cell type/ marker gene combination
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- all_info_df |>
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  dplyr::summarize(
    .by = c("label_recoded", "ensembl_gene_id"),
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |>
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments
  dplyr::left_join(nbatlas_markers_df, by = c("ensembl_gene_id", "label_recoded"), relationship = "many-to-many") |> # add total cells
  dplyr::left_join(total_cells_df, by = c("label_recoded")) |>
  # only show cells that have above the minimum total number
  dplyr::filter(total_cells > min_cells) |>
  dplyr::mutate(
    # get total percent expressed
    percent_exp = (detected_count / total_cells) * 100,
    # account for NA/unknowns and set axes order
    broad_celltype_group = factor(label_recoded, levels = c(names(celltype_pal)))
  )

# specify x axis order for dotplot
marker_gene_order <- nbatlas_markers_df |>
  dplyr::pull(gene_symbol) |>
  unique()

# set order for cell types
celltype_order <- unique(nbatlas_markers_df$label_recoded)

# filter out low expressed genes
dotplot_df <- group_stats_df |>
  dplyr::filter(mean_exp > 0, percent_exp > 10) |>
  dplyr::arrange(label_recoded) |>
  # add a label for the plot
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})")) |>
  # remove marker genes that aren't present in final annotations and set x axis order
  dplyr::filter(gene_symbol %in% marker_gene_order) |>
  dplyr::mutate(
    # set orders of gene symbol and validation groups
    y_label = factor(y_label, levels = rev(unique(y_label))),
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    label_recoded = factor(label_recoded, levels = celltype_order)
  )

# make dotplot with marker gene exp
dotplot <- ggplot(dotplot_df, aes(y = y_label, x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "transparent", color = NA),
    strip.placement = "outside",
    strip.text.x = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.ticks.x = element_blank(),
    text = element_text(size = 14),
    panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
  ) +
  labs(
    x = "",
    y = "label_recoded",
    color = "Mean gene expression",
    size = "Percent cells expressed"
  )

# add annotation bar aligning marker genes with validation group
color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = label_recoded)) +
  geom_tile() +
  facet_grid(cols = vars(label_recoded), scales = "free", space = "free") +
  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$label_recoded)) +
  ggmap::theme_nothing() +
  theme(
    strip.background = element_rect(fill = "transparent", color = NA),
    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
    strip.placement = "outside",
    legend.position = "none",
    panel.spacing = unit(0.5, "lines"),
    strip.clip = "off"
  ) +
  labs(fill = "")

>>>>

```{r}
# get top 10??? upregulated marker genes per cell type
nbatlas_markers_df <- readr::read_tsv("../references/nbatlas-marker-genes.tsv") |>
  dplyr::filter(direction == "up") |>
  dplyr::group_by(NBAtlas_label) |>
  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
  dplyr::ungroup() |>
  dplyr::filter(rank_lfc <= 10) |>
  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
  dplyr::select(NBAtlas_label_recoded, gene_symbol, ensembl_gene_id, avg_log2FC)
```

@@

# get markers that are expressed
expressed_markers <- intersect(all_markers, expressed_genes)
```

@@

# Join with cell type results and marker gene info
all_info_df <- celltype_recoded_df |>
  dplyr::filter(
    annotation_type == "singler",
    # we don't consider the NAs here
    label_recoded != "singler_unknown"
  ) |>
  dplyr::select(cell_id, label_recoded) |>
  dplyr::left_join(gene_exp_df, by = "cell_id") |>
  # account for the same gene being present in multiple cell types
  dplyr::left_join(nbatlas_markers_df, by = "ensembl_gene_id", relationship = "many-to-many")
```

```{r}
min_cells <- 0

# prep for plots
# get total number of cells per final annotation group
total_cells_df <- all_info_df |>
  dplyr::select(cell_id, label_recoded) |>
  dplyr::distinct() |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::arrange(desc(total_cells)) |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))

# y axis - singler
y_label_order <- rev(total_cells_df$y_label)

# bar order - NBAtlas
nbatlas_order <- total_cells_df$label_recoded

# x axis marker genes - nbatlas
nbatlas_markers_df <- nbatlas_markers_df |>
  dplyr::filter(ensembl_gene_id %in% expressed_markers) |>
  dplyr::mutate(
    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
  ) |>
  dplyr::arrange(NBAtlas_label_recoded)

marker_gene_order <- nbatlas_markers_df |>
  dplyr::pull(gene_symbol) |>
  unique()

# table with one row per unique broad cell type/ marker gene combination
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- all_info_df |>
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  # remove genes that aren't present in final annotations
  dplyr::filter(gene_symbol %in% marker_gene_order) |>
  dplyr::group_by(label_recoded, NBAtlas_label_recoded, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |>
  dplyr::ungroup() |>
  # add total cells
  dplyr::left_join(total_cells_df, by = "label_recoded") |>
  # get total percent expressed
  dplyr::mutate(percent_exp = (detected_count / total_cells) * 100) |>
  # remove lowly expressed marker genes and any that aren't present in final annotations
  dplyr::filter(
    mean_exp > 0,
    percent_exp > 10,
    # only show cells that have above the minimum total number
    total_cells > min_cells
  ) |>
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments
  dplyr::left_join(
    nbatlas_markers_df,
    by = c("ensembl_gene_id", "NBAtlas_label_recoded"),
    relationship = "many-to-many"
  )

dotplot_df <- group_stats_df |>
  dplyr::mutate(
    # set orders of gene symbol and validation groups
    y_label = factor(y_label, y_label_order),
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    NBAtlas_label_recoded = factor(NBAtlas_label_recoded, levels = nbatlas_order)
  )
```

```{r}
# make dotplot with marker gene exp
dotplot <- ggplot(dotplot_df, aes(y = y_label, x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "transparent", color = NA),
    strip.placement = "outside",
    strip.text.x = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.ticks.x = element_blank(),
    text = element_text(size = 14),
    panel.spacing = unit(0.5, "lines") # adjust spacing and match with annotation bar
  ) +
  labs(
    x = "NBAtlas cell type marker genes",
    y = "SingleR label",
    color = "Mean gene expression",
    size = "Percent cells expressed"
  )

# add annotation bar aligning marker genes with validation group
color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = NBAtlas_label_recoded)) +
  geom_tile() +
  facet_grid(cols = vars(NBAtlas_label_recoded), scales = "free", space = "free") +
  scale_fill_manual(values = celltype_pal, breaks = levels(dotplot_df$NBAtlas_label_recoded)) +
  ggmap::theme_nothing() +
  theme(
    strip.background = element_rect(fill = "transparent", color = NA),
    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
    strip.placement = "outside",
    legend.position = "none",
    panel.spacing = unit(0.5, "lines"),
    strip.clip = "off"
  ) +
  labs(fill = "")
